A non-contextual hidden variable model for quantum mechanics

### 2016-07-11 preReleases/version6/finalPaper.tex
added the PDFs plus released version 6

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}
  \theoremstyle{definition}
  \newtheorem{defn}{\protect\definitionname}

@@

  \theoremstyle{plain}
  \newtheorem*{prop*}{\protect\propositionname}
  \theoremstyle{plain}
  \newtheorem{lem}[thm]{\protect\lemmaname}
  \theoremstyle{plain}
  \newtheorem{prop}[thm]{\protect\propositionname}
  \theoremstyle{definition}
  \newtheorem{example}[thm]{\protect\examplename}
  \theoremstyle{definition}
  \newtheorem{notn}[thm]{\protect\notationname}

@@

{\noindent \bf Notation:} (a) $\psi\in\mathcal{H}$ 
represents a pure quantum mechanical
state of the system in the Hilbert space
$\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined
to mean the set of Hermitian 
observables for the system, (c)
$[\mathcal{H}]$ is defined to mean
$(\mathcal{H},\,\mathbb{R}^{\otimes})$, which
represents the state of the system including HVs,
(d) $[\psi]\in[\mathcal{H}]$ will represent the
state of the system, including HVs, (e) a
prediction map is
$M:\{ \hat{\mathcal{H}},[\mathcal{H}] \}\to\mathbb{R}$,
(f) a sequence map is
$S:\{ \hat{\mathcal{H}},[\mathcal{H}],\mathbb{R} \}\to[\mathcal{H}]$,
(g) $f$ is an arbitrary map from $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\} \to \hat{\mathcal{H}}$ constructed using
multiplication and addition of the observables,
and multiplication with complex
numbers\footnote{Strictly, the map $f$ would
depend on the operators to ensure preservation of
Hermiticity; thus a given $f$ will be defined only
for a subset of $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\}$.}, (h) $\tilde{f}$ is a map constructed by
replacing observables in $f$ with real numbers.

@@

Before we re-examine the result of Kochen and
Specker~\cite{KochenSpecker} that non-contextual
models cannot be compatible with quantum
mechanics and show that multiplicativity can provide an
alternative to contextuality, we define
multiplicativity more precisely.
\begin{defn}
A prediction map $M$ is \emph{multiplicative} iff
\begin{align*} &
M(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi])
= \\ & \tilde
f(M(\hat{B}_{1},[\psi]),M(\hat{B}_{2},[\psi]),\dots
M(\hat{B}_{N},[\psi])), \end{align*}
where $\hat{B}_{i}\in\hat{\mathcal{H}}$ are
arbitrary mutually commuting observables and
$[\psi]\in[\mathcal{H}]$. A
\emph{non-multiplicative} map is one that is not
multiplicative.  
\end{defn} 
Note that if $M$ is taken to represent the
measurement outcome (in QM), then for states of
the system which are simultaneous eigenkets of
$\hat{B}_{i}$s, $M$ must clearly be
multiplicative. It is, however, not obvious that
this property must always hold. For example,
consider two spin-half particles in the state 
$\left|1\right\rangle \otimes\left|1\right\rangle
$ written in the computational basis and take
operators 
$\hat{B}_{1}=\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}$,
$\hat{B}_{2}=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}$
and
$\hat{C}=\hat{B}_{1}\hat{B}_{2}=-\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$
written in terms of Pauli operators.
We must have $M(\hat{C})=-1$ while
$M(\hat{B}_{1})=\pm1$ and $M(\hat{B}_{2})=\pm1$
independently, according to QM, with probability
half. Here multiplicativity does not seem to hold.
Antithetically, it is clear that if one first
measures $\hat{B}_{1}$ and subsequently measures
$\hat{B}_{2}$, then the product of the results
must be $-1$. This is consistent with measuring
$\hat{C}$. To capture this property we define
\emph{sequential multiplicativity} as follows.
\begin{defn} 
A prediction map $m$ is
\emph{sequentially multiplicative} for a given
sequence map $s$, iff \begin{align*}
&M(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi_{1}])=\\
&\tilde
f(M(\hat{B}_{1},[\psi_{k_{1}}]),M(\hat{B}_{2},[\psi_{k_{2}}]),\dots,M(\hat{B}_{N},[\psi_{k_{N}}])),
\end{align*} 
where
$\mathbf{k}=(k_{1},k_{2},\dots,k_{N})\in\{
N! {\rm ~permutations~of~}  k'{\rm s}
\}$, $\hat{B}_{i}\in\hat{\mathcal{H}}$ are
arbitrary mutually commuting observables,
$[\psi_{i}]\in[\mathcal{H}]$ and
$[\psi_{k+1}] :=
S(\hat{B}_{k},[\psi_{k}],M(\hat{B}_{k},[\psi_{k}]))$,
$\forall\,[\psi_{i}]$.  
\label{defn:seqnMltpl}\end{defn} 
With these definitions we are now ready to discuss
the `proof of contextuality'. We first state the
contextuality theorem in our notation:
\begin{thm} Let a map
$M:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a)
$M(\hat{\mathbb{I}})=1$, (b)
$M(f(\hat{B}_{1},\hat{B}_{2},\dots))=\tilde
f(M(\hat{B}_{1}),M(\hat{B}_{2}),\dots)$, for any
arbitrary function $f$, where $\hat{B}_{i}$ are
mutually commuting Hermitian operators. If $m$ is
assumed to describe the outcomes of measurements,
then no $M$ exists which is consistent with all
predictions of QM. 
\label{thm:KS}
\end{thm}

\begin{proof} Peres Mermin
($\left|\mathcal{H}\right|\ge4$)
\cite{Peres,Mermin}:
For a system composed of two spin-half particles 
consider the following set of operators \[
\hat{A}_{ij}\doteq\left[\begin{array}{ccc}
\hat{\mathbb{I}}\otimes\hat{\sigma}_{x} &
\hat{\sigma}_{x}\otimes\hat{\mathbb{I}} &
\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}\\
\hat{\sigma}_{y}\otimes\hat{\mathbb{I}} &
\hat{\mathbb{I}}\otimes\hat{\sigma}_{y} &
\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}\\
\hat{\sigma}_{y}\otimes\hat{\sigma}_{x} &
\hat{\sigma}_{x}\otimes\hat{\sigma}_{y} &
\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}
\end{array}\right] \] which have the property that
all operators along a row (or column) commute. Further,
the product of rows (or columns) yields
$\hat{R}_{i}=\mathbb{I}$ and
$\hat{C}_{j}=\mathbb{I}\,(j\neq3)$,
$\hat{C}_{3}=-\mathbb{I}$, ($\forall\,i,j$) where
$\hat{R}_{i}:=\prod_{j}\hat{A}_{ij}$,
$\hat{C}_{j}:=\prod_{i}\hat{A}_{ij}$. Let us
assume that $M$ exists. From property (b) of the
map, to get $M(\hat{C}_{3})=-1$ (as required by
property (a)), we must have an odd number of $-1$
assignments in the third column. In the remaining
columns, the number of $-1$ assignments must be
even (for each column). Thus, in the entire
square, the number of $-1$ assignments must be
odd. Let us use the same reasoning, but along the
rows. Since each $M(\hat{R}_{i})=1$, we must have
even number of $-1$ assignments along each row.
Thus, in the entire square, the number of $-1$
assignments must be even. We have arrived at a
contradiction and therefore we conclude that our
assumption that a consistent $M$ exists, must be
wrong.  \end{proof}

\begin{thmrem} One could in principle assume $M$, to
be s.t. (a) $M(\hat{\mathbb{I}})=1$, (b)
$M(\alpha\hat{B}_{i})=\alpha M(\hat{B}_{i})$, for
$\alpha\in\mathbb{R}$, (c)
$M(\hat{B}_{i}^{2})=M(\hat{B}_{i})^{2}$, (d)
$M(\hat{B}_{i}+\hat{B}_{j})=M(\hat{B}_{i})+M(\hat{B}_{j})$,
to deduce (d')
$M(\hat{B}_{i}\hat{B}_{j})=M(\hat{B}_{i})M(\hat{B}_{j})$
and that $M(\hat{B}_{i})\in$ spectrum of
$\hat{B}_{i}$. Effectively then, condition (b)
listed in the theorem is satisfied as a
consequence.  Therefore, assuming (a)-(d) as
listed here, rules out a larger class of $M$.
\cite{KochenSpecker} 
\end{thmrem} 
Here $M$ maybe viewed as a specific
class of prediction maps, that implicitly depends
on the state $[\psi]$. It is clear that according to
\thmref{KS}, non-contextual maps which are
\emph{multiplicative} must be incompatible with
QM. However, non-contextual maps which are
non-\emph{multiplicative} might be consistent with
QM. Indeed it is and we will show this by constructing an
explicit model.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}

\theoremstyle{plain}
\newtheorem*{thm*}{\protect\theoremname}

  \theoremstyle{definition}
  \newtheorem{defn}{\protect\definitionname}

@@

  \theoremstyle{plain}
  \newtheorem*{prop*}{\protect\propositionname}
  \theoremstyle{plain}
  \newtheorem{lem}[thm]{\protect\lemmaname}
  \theoremstyle{plain}
  \newtheorem{prop}[thm]{\protect\propositionname}
  \theoremstyle{definition}
  \newtheorem{example}[thm]{\protect\examplename}
  \theoremstyle{definition}
  \newtheorem*{notn*}{\protect\notationname}

@@

%{\noindent \bf Notation:} 
\begin{notn*}
(a) $\psi\in\mathcal{H}$ 
represents a pure quantum mechanical
state of the system in the Hilbert space
$\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined
to mean the set of Hermitian 
observables for the system, (c)
$[\mathcal{H}]$ is defined to mean
$(\mathcal{H},\,\mathbb{R}^{\otimes})$, which
represents the state of the system including HVs,
(d) $[\psi]\in[\mathcal{H}]$ will represent the
state of the system, including HVs, (e) a
prediction map is
$M:\{ \hat{\mathcal{H}},[\mathcal{H}] \}\to\mathbb{R}$,
(f) a sequence map is
$S:\{ \hat{\mathcal{H}},[\mathcal{H}],\mathbb{R} \}\to[\mathcal{H}]$,
(g) $f$ is an arbitrary map from $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\} \to \hat{\mathcal{H}}$ constructed using
multiplication and addition of the observables,
and multiplication with complex
numbers\footnote{Strictly, the map $f$ would
depend on the operators to ensure preservation of
Hermiticity; thus a given $f$ will be defined only
for a subset of $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\}$.}, (h) $\tilde{f}$ is a map constructed by
replacing observables in $f$ with real numbers.
\end{notn*}

@@

Before we re-examine the result of Kochen and
Specker~\cite{KochenSpecker} that non-contextual
models cannot be compatible with quantum
mechanics and show that multiplicativity can provide an
alternative to contextuality, we define
multiplicativity more precisely.
\begin{defn}
A prediction map $M$ is \emph{multiplicative} iff
\begin{align*} &
M(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi])
= \\ & \tilde
f(M(\hat{B}_{1},[\psi]),M(\hat{B}_{2},[\psi]),\dots
M(\hat{B}_{N},[\psi])), \end{align*}
where $\hat{B}_{i}\in\hat{\mathcal{H}}$ are
arbitrary mutually commuting observables and
$[\psi]\in[\mathcal{H}]$. A
\emph{non-multiplicative} map is one that is not
multiplicative.  
\end{defn} 
Note that if $M$ is taken to represent the
measurement outcome (in QM), then for states of
the system which are simultaneous eigenkets of
$\hat{B}_{i}$s, $M$ must clearly be
multiplicative. It is, however, not obvious that
this property must always hold. For example,
consider two spin-half particles in the state 
$\left|1\right\rangle \otimes\left|1\right\rangle
$ written in the computational basis and take
operators 
$\hat{B}_{1}=\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}$,
$\hat{B}_{2}=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}$
and
$\hat{C}=\hat{B}_{1}\hat{B}_{2}=-\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$
written in terms of Pauli operators.
We must have $M(\hat{C})=-1$ while
$M(\hat{B}_{1})=\pm1$ and $M(\hat{B}_{2})=\pm1$
independently, according to QM, with probability
half. Here multiplicativity does not seem to hold.
Antithetically, it is clear that if one first
measures $\hat{B}_{1}$ and subsequently measures
$\hat{B}_{2}$, then the product of the results
must be $-1$. This is consistent with measuring
$\hat{C}$. To capture this property we define
\emph{sequential multiplicativity} as follows.
\begin{defn} 
A prediction map $m$ is
\emph{sequentially multiplicative} for a given
sequence map $s$, iff \begin{align*}
&M(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi_{1}])=\\
&\tilde
f(M(\hat{B}_{1},[\psi_{k_{1}}]),M(\hat{B}_{2},[\psi_{k_{2}}]),\dots,M(\hat{B}_{N},[\psi_{k_{N}}])),
\end{align*} 
where
$\mathbf{k}=(k_{1},k_{2},\dots,k_{N})\in\{
N! {\rm ~permutations~of~}  k'{\rm s}
\}$, $\hat{B}_{i}\in\hat{\mathcal{H}}$ are
arbitrary mutually commuting observables,
$[\psi_{i}]\in[\mathcal{H}]$ and
$[\psi_{k+1}] :=
S(\hat{B}_{k},[\psi_{k}],M(\hat{B}_{k},[\psi_{k}]))$,
$\forall\,[\psi_{i}]$.  
\label{defn:seqnMltpl}\end{defn} 
With these definitions we are now ready to discuss
the `proof of contextuality'. We first state the
contextuality theorem in our notation:
\begin{thm*} Let a map
$M:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a)
$M(\hat{\mathbb{I}})=1$, (b)
$M(f(\hat{B}_{1},\hat{B}_{2},\dots))=\tilde
f(M(\hat{B}_{1}),M(\hat{B}_{2}),\dots)$, for any
arbitrary function $f$, where $\hat{B}_{i}$ are
mutually commuting Hermitian operators. If $m$ is
assumed to describe the outcomes of measurements,
then no $M$ exists which is consistent with all
predictions of QM. 
\label{thm:KS}
\end{thm*}

\begin{proof} Peres Mermin [PM]
($\left|\mathcal{H}\right|\ge4$)
\cite{Peres,Mermin}:
For a system composed of two spin-half particles 
consider the following set of operators \[
\hat{A}_{ij}\doteq\left[\begin{array}{ccc}
\hat{\mathbb{I}}\otimes\hat{\sigma}_{x} &
\hat{\sigma}_{x}\otimes\hat{\mathbb{I}} &
\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}\\
\hat{\sigma}_{y}\otimes\hat{\mathbb{I}} &
\hat{\mathbb{I}}\otimes\hat{\sigma}_{y} &
\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}\\
\hat{\sigma}_{y}\otimes\hat{\sigma}_{x} &
\hat{\sigma}_{x}\otimes\hat{\sigma}_{y} &
\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}
\end{array}\right] \] which have the property that
all operators along a row (or column) commute. Further,
the product of rows (or columns) yields
$\hat{R}_{i}=\mathbb{I}$ and
$\hat{C}_{j}=\mathbb{I}\,(j\neq3)$,
$\hat{C}_{3}=-\mathbb{I}$, ($\forall\,i,j$) where
$\hat{R}_{i}:=\prod_{j}\hat{A}_{ij}$,
$\hat{C}_{j}:=\prod_{i}\hat{A}_{ij}$. Let us
assume that $M$ exists. From property (b) of the
map, to get $M(\hat{C}_{3})=-1$ (as required by
property (a)), we must have an odd number of $-1$
assignments in the third column. In the remaining
columns, the number of $-1$ assignments must be
even (for each column). Thus, in the entire
square, the number of $-1$ assignments must be
odd. Let us use the same reasoning, but along the
rows. Since each $M(\hat{R}_{i})=1$, we must have
even number of $-1$ assignments along each row.
Thus, in the entire square, the number of $-1$
assignments must be even. We have arrived at a
contradiction and therefore we conclude that our
assumption that a consistent $M$ exists, must be
wrong.  \end{proof}

\begin{thmrem} One could in principle assume $M$, to
be s.t. (a) $M(\hat{\mathbb{I}})=1$, (b)
$M(\alpha\hat{B}_{i})=\alpha M(\hat{B}_{i})$, for
$\alpha\in\mathbb{R}$, (c)
$M(\hat{B}_{i}^{2})=M(\hat{B}_{i})^{2}$, (d)
$M(\hat{B}_{i}+\hat{B}_{j})=M(\hat{B}_{i})+M(\hat{B}_{j})$,
to deduce (d')
$M(\hat{B}_{i}\hat{B}_{j})=M(\hat{B}_{i})M(\hat{B}_{j})$
and that $M(\hat{B}_{i})\in$ spectrum of
$\hat{B}_{i}$. Effectively then, condition (b)
listed in the theorem is satisfied as a
consequence.  Therefore, assuming (a)-(d) as
listed here, rules out a larger class of $M$.
\cite{KochenSpecker} 
\end{thmrem} 
Here $M$ maybe viewed as a specific
class of prediction maps, that implicitly depends
on the state $[\psi]$. It is clear that according to
the theorem, non-contextual maps which are
\emph{multiplicative} must be incompatible with
QM. However, non-contextual maps which are
non-\emph{multiplicative} might be consistent with
QM. Indeed it is and we will show this by constructing an
explicit model.
